---
title: "Cell-cell_communication analysis"
output: html_document
---

# Introduction to cell-cell communication analysis

*link to powerpoint slides:* A power point presentation will describe the general principles of single-cell 'omics' and basics about the platforms (\~15 - 20 minutes)

------------------------------------------------------------------------

Cells communicate through various mechanisms including chemical signals, physical properties (e.g., mechanotransduction), secreted proteins, and other protein-protein interactions.

![Generated using Biorender.com](images/Untitled%20(7)-01.png){width="586"}

The analysis of cell-cell signaling from single-cell transcriptomic is based on the broad understandings/assumption that [genes encode both ligands and receptors,]{.underline} and sometimes, that activation of a signaling pathway drives and transcriptomic response, both of which can be detected as mRNA expression in individual cells.

------------------------------------------------------------------------

## 0. Setting up the R environment

```{r}
suppressPackageStartupMessages({
  library(CellChat)  
  library(nichenetr)
  library(Seurat)
})

```

**Download pre-processed data for liver from mice treated with TCDD**

```{r}
options(timeout = max(6000, getOption("timeout")))
download.file('https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE148339&format=file&file=GSE148339%5Fliver%2Eintegrated%2Efinal%2ERData%2Egz', './downloads/GSE148339_liver.integrated.final.RData.gz')
system("gunzip ./downloads/GSE148339_liver.integrated.final.RData.gz")
```

**Load the downloaded object**

```{r}
liver_seurat_data = readRDS('./downloads/GSE148339_liver.integrated.final.RData')
liver_seurat_data = UpdateSeuratObject(liver_seurat_data)
```

Next we will take a look at the data to make sure it is loaded properly and that we understand the dataset.

```{r}
DimPlot(liver_seurat_data, group.by = 'celltype', label = TRUE) + NoLegend()
```

For this workshop we will focus on a smaller subset of the data (only endothelial cells and stellate cells).

```{r}
endo_stellate = subset(liver_seurat_data, idents = c("Endothelial Cells", "Stellate Cells"))
nonzero_rows <- which(rowSums(as.matrix(endo_stellate@assays$RNA@counts)) != 0)
endo_stellate = endo_stellate[nonzero_rows, ]
endo_stellate@meta.data$celltype = factor(endo_stellate@meta.data$celltype, levels=c('Endothelial Cells', 'Stellate Cells'))
```

## 1. Determining *known* cell-cell communication signals

Cell-cell communication analysis tools rely on *a priori* knowledge of interactions. This means that they need to have access to reliable data which represents true and biologically meaningful interactions. Therefore, many of these tools obtain their information from shared source (*e.g.*, KEGG, Reactome, STRING). Nevertheless, there is a lot of heterogeneity in the interactions which are captured across various tools which is explored in further depth by [Dimitrov et al. (2024)](https://www.nature.com/articles/s41467-022-30755-0).

![Similarity between the different resources based on the interactions (Jaccard Index) from Dimitrov, D., Türei, D., Garrido-Rodriguez, M. et al. Comparison of methods and resources for cell-cell communication inference from single-cell RNA-Seq data. Nat Commun 13, 3224 (2022). <https://doi.org/10.1038/s41467-022-30755-0>](images/clipboard-747365050.png)

### 1.1. Databases and knowledgebases

As was just highlighted, cell-cell communication analysis is dependent on the underlying database/knowledgebase of existing interactions. Lets take a look at some of the protein-protein interactions which are included in the CellChat database, one of the many commonly use cell-cell communication analysis tools.

```{r}
print(head(CellChatDB.human$interaction))
```

This is an example of how these tools curate interactions for subsequent analyses, identifying the ligand(s), receptor(s), type of signaling, cellular location, and gene symbols. However, its important to note that this database is based on [*known human interactions*]{.underline}.

```{r}
print(paste0(
'Size of human database: ',
nrow(CellChatDB.human$interaction)
))

print(paste0(
'Size of mouse database: ',
nrow(CellChatDB.mouse$interaction)
))

print(paste0(
'Size of zebrafish database: ',
nrow(CellChatDB.zebrafish$interaction)
))
```

We can also look at tools such as NicheNet, which include not only cell-cell interactions, but the subsequent signaling cascade as well. We can load the prior knowledge (PK) as follows:

```{r}
# Load PK
ligand_target_matrix <- readRDS(url("https://zenodo.org/record/7074291/files/ligand_target_matrix_nsga2r_final.rds"))
lr_network <- readRDS(url("https://zenodo.org/record/7074291/files/lr_network_human_21122021.rds"))
```

Next we can examine the database of ligand-receptor interactions. We will filter for TGFB1 as the ligand, just for simplicity.

```{r}
lr_network %>% filter(from == "TGFB1")
```

We can see a structure that is very similar to CellChat. However, NicheNet relies on a second piece of information - the probabilites that certain genes are induced in response to activation of a specific receptor. Here is TGFBR2 for example:

```{r}
ligand_target_matrix['TGFBR2', 1:4]
```

## 2. Steady-state and differential *inter*cellular communication

As with any analysis, whether its in the lab or purely computational, it is important to know what is the right tool to answer your question/test your hypothesis. *Are you looking to characterize cell-cell communication in a tissue without any treatment or perturbation, or are you looking to test the effect of a perturbation on cell-cell communication?*

### 2.1. Steady state cell-cell communication

Expression based cell-cell communication analysis tools, including the popular [CellPhoneDB](https://doi.org/10.1038/s41596-020-0292-x) and [CellChat](https://doi.org/10.1038/s41467-021-21246-9) ools, were primarily developed to infer ligand-receptor interactions within cells that are at a steady state. Tools such as CellChat have also developed condition comparison methods. These will be discussed later.

For the exact algorithms which each of these tools use, someone should always review the original publication. [*However, many of these tools share a similar overall concept which we will illustrate using the code below.*]{.underline}

```{r}
# Load required libraries
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Generate simulated data for interactionA and interactionB
n_samples <- 100

# Interaction A: Samples from two populations with very different means
interactionA_pop1 <- rnorm(n_samples, mean = 5, sd = 1)
interactionA_pop2 <- rnorm(n_samples, mean = 10, sd = 1)
interactionA <- c(interactionA_pop1, interactionA_pop2)

# Interaction B: Samples from two populations with the same means
interactionB_pop1 <- rnorm(n_samples, mean = 15, sd = 2)
interactionB_pop2 <- rnorm(n_samples, mean = 15, sd = 3)
interactionB <- c(interactionB_pop1, interactionB_pop2)

# Function to shuffle data
shuffle_data <- function(data) {
  shuffled_data <- sample(data)
  return(shuffled_data)
}

# Function to create violin plot
create_violin_plot <- function(original_data, shuffled_data, interaction_name) {
  original_df <- data.frame(Group = rep(c("Pop1", "Pop2"), each = n_samples),
                            Value = c(original_data, shuffled_data),
                            Type = rep(c("Original", "Shuffled"), each = length(original_data)))
  
  ggplot(original_df, aes(x = Group, y = Value, fill = Type)) +
    geom_violin() +
    labs(title = interaction_name, x = "Group", y = "Value", fill = "Type") +
    theme_minimal()
}

# Plot violin plots for interactionA
violin_plot_interactionA <- create_violin_plot(interactionA, shuffle_data(interactionA), "Interaction A")

# Plot violin plots for interactionB
violin_plot_interactionB <- create_violin_plot(interactionB, shuffle_data(interactionB), "Interaction B")

# Plot violin plots
gridExtra::grid.arrange(violin_plot_interactionA, violin_plot_interactionB, nrow = 1)



```

Note how interaction A shows clear differences in the original data which is lost after shuffling. Alternatively, interaction B is similar whether it is shuffled or not. In this illustrative example, *interaction A* would likely have been considered significant while *interaction B* would not have been even though it is more highly expressed.

#### 2.1a. CellChat example

```{r}
cellChat <- createCellChat(object = endo_stellate, group.by = "celltype", assay = "RNA")
CellChatDB.use <- CellChatDB.mouse
cellchat@DB <- CellChatDB.use
```

```{r}
# subset the expression data of signaling genes for saving computation cost
cellchat <- subsetData(cellchat) # This step is necessary even if using the whole database
cellchat <- identifyOverExpressedGenes(cellchat)
cellchat <- identifyOverExpressedInteractions(cellchat)
```

```{r}
cellchat <- computeCommunProb(cellchat, type = "triMean")
cellchat <- filterCommunication(cellchat, min.cells = 10)
cellchat <- computeCommunProbPathway(cellchat)
cellchat <- aggregateNet(cellchat)
```

```{r}
groupSize <- as.numeric(table(cellchat@idents))

par(mfrow = c(1,2), xpd=TRUE)
netVisual_circle(cellchat@net$count, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Number of interactions")
netVisual_circle(cellchat@net$weight, vertex.weight = groupSize, weight.scale = T, label.edge= F, title.name = "Interaction weights/strength")
```

### 2.2. Perturbation of cell-cell communication

#### 2.2a. Enrichment-like approach (NicheNet)

[NicheNet](https://github.com/saeyslab/nichenetr) is unique among cell-cell communication in the sense that it does require expression data as input, and it considers downstream effectors such as transcription factors and responses.

Let's consider the question; what ligand-receptor interactions are induced by treatment with chemical X in multiple immune cell types (*e.g.*, T Cells, B Cells, and Macrophages) coming from a source cell type (*e.g.*, hepatocytes). Here we would obtain the following inputs:

1.  A list of ligands expressed in source/sender cells and a list of receptors in the receiving cells. This list is used to refine the interactions of interest to those that are possible based on gene expression

    ```{r}
    ligands_expressed = intersect(toupper(rownames(endo_stellate)), lr_network$from) 
    receptors_expressed = intersect(toupper(rownames(endo_stellate)), lr_network$to)
      
    potential_interactions <- lr_network %>% 
      filter(from %in% ligands_expressed & to %in% receptors_expressed) %>%
      pull(from) %>% unique()
    ```

2.  A list of all genes expressed in the receiver cell type:

    ```{r}
    background_genes = toupper(rownames(endo_stellate))
    ```

3.  A list of genes which are induced from a treatment or condition:

    ```{r}
    DEAnalysis = FindMarkers(endo_stellate, ident.1 = 'Endothelial Cells', ident.2 = 'Stellate Cells', min.pct = 0.5, logfc.treshold = log(2), min.diff.pct = 0.25)
    geneset_oi = toupper(rownames(DEAnalysis %>% filter(avg_log2FC > 0))) # geneset of interest
    ```

Next we run the analysis of ligand activity and output the top 10 inferred active ligands:

```{r}
ligand_activities <- predict_ligand_activities(
  geneset = geneset_oi, 
  background_expressed_genes = background_genes,
  ligand_target_matrix = ligand_target_matrix,
  potential_ligands = ligands_expressed)

ligand_activities <- ligand_activities %>% 
  arrange(-aupr) %>% 
  mutate(rank = rank(desc(aupr)))

# show top10 ligand activities
head(ligand_activities, n=10)
```

#### 2.2b. CellChat

```{r}

control_seurat <- subset(endo_stellate, subset = treatment == "CONTROL")
treated_seurat <- subset(endo_stellate, subset = treatment == "TCDD")

cellChat_control <- createCellChat(object = control_seurat, group.by = "celltype", assay = "RNA")
cellChat_treated <- createCellChat(object = treated_seurat, group.by = "celltype", assay = "RNA")

object.list <- list(CONTROL = cellChat_control, TCDD = cellChat_treated)

for (cc in names(object.list)){
  cellchat = object.list[[cc]]
  CellChatDB.use = CellChatDB.mouse
  cellchat@DB = CellChatDB.use
  cellchat = subsetData(cellchat)
  cellchat = identifyOverExpressedGenes(cellchat)
  cellchat = identifyOverExpressedInteractions(cellchat)
  cellchat = computeCommunProb(cellchat, type = "triMean")
  cellchat = filterCommunication(cellchat, min.cells = 10)
  cellchat = computeCommunProbPathway(cellchat)
  cellchat = aggregateNet(cellchat)
  object.list[[cc]] = cellchat
  
}

cellchat <- mergeCellChat(object.list, add.names = names(object.list))

```

```{r}
gg1 <- rankNet(cellchat, mode = "comparison", measure = "weight", sources.use = NULL, targets.use = NULL, stacked = T, do.stat = TRUE)
gg2 <- rankNet(cellchat, mode = "comparison", measure = "weight", sources.use = NULL, targets.use = NULL, stacked = F, do.stat = TRUE)

gg1 + gg2
```

## 2. Interpretation and visualization

### Choosing the types of interactions of interest

### Another thing

### Another thing 2

## 3. Advanced approaches
